Research Experience

### 2010-03-23 doc/thesis/thesis.tex
more particle work.

\subsection*{Introduction to Particle Filters}
Particle filters, a type of Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, Sequential Monte-Carlo methods are designed
to be used with parameters that vary with time. Unlike variations of the
Kalman filter, particle filters do not make the assumption that noise
is Gaussian. Thus particle filters are often the best solution to bayesian 
tracking for non-linear, non-gaussian systems. 
%todo: might need to reference attempts to model a system of our type with
%unscented kalman filter
The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF obviously introduces bias; however, every gradient
descent must choose starting points and it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose a set or stream of measurements
are given, $\{y(t), t = 1, 2, 3, ... T\}$, where $T$ is permitted to go
 to infinity. Then the goal is to find the 
parameters, $\hat{\theta}$, and underlying state time series, $\hat{x}[0:T]$
that minimize the difference between $\hat{y}[0:T]$
and $y[0:T]$. In our case, we will assume that we know the form
of the model, which is based on first principals, and that
there is some true $\theta$ and a true time-series of underlying
state variable, $x[0:T]$ that drives $y[0:T]$. Assuming a model form 
such as we do here reduces model variance, potentially at the cost of increased
bias (or systematic) error. We will assume a basic state space model:

@@

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, the systems are assumed to be time invariant. This 
assumption is based on the idea that if you froze the system for $\delta t$
seconds, when unfrozen the system would continue as if nothing happend. 
Few biological systems are predictible enough for them to be summarized
by a time varying function. Although the heart may seem like an obvious
exception, period between heartbeats vary often enough that prediction
would necessate another state-space model. In short, we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Third, we will assume noise is additive.
Finally, $x(t)$ will encapsulate $\theta$, the unknown model constants, which
means that the vector $\dot{x}$ will always have members
that are 0. The results of these assumptions are a simplified version of the
state space equations:

@@

$\nu_x$ will result in something akin to adding a weiner process to $y[0:T]$, and thus
will include low frequency noise. $\nu_y$ on the other hand will cause i.i.d. noise
in $y[0:T]$. For some tests I will use de-trending methods to reduce the effects of 
$\nu_x$, the remainder of which gets re-factored into $\nu_y$. Both $\nu_x$ and $\nu_y$
have biological and non-biological sources. MR can lead to both types of noise, 
as demonstrated in \cite{drift}. Meanwhile changes in metabolism, heart rate, or
other biochemical intervention could all lead to either $\nu_x$ or $\nu_y$.

The goal of the particle filter to evolve a probably distribution $\hat{x}(T)$,
that asymptotically approaches the probability distribution of $x(t)$,
given known inputs, $u[0:T]$ and noisy readings $y[0:T]$.
To begin with, the particle filter starts with a prior distribution,
which is used to build a Dirac Mixture Probability Distribution Function (PDF):

\begin{equation}
\hat{x}(0) \sim p(X = x) = \frac{1}{n}\sum_{i = 0}^{n-1}p_i(x)
\end{equation}
where $p_i(x)$ is some probability density function centered at $\hat{x}_i(0)$, particle
$i$ at time $0$. Thus the sum of the density at $x$ from each particles' PDF is the density
at $x$. If
the distributions are infinitely thin, then the distribution will be discrete, otherwise
it will be continuous.

>>>>

\subsection*{Introduction to Particle Filters}
Particle filters, a type of Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters give a timeseries of measurements. Unlike Markov 
Chain Monte Carlo estimation, Sequential Monte-Carlo methods are designed
to be used with parameters that vary with time. Unlike variations of the
Kalman filter, particle filters do not make the assumption that noise
is Gaussian. Thus particle filters are often the best solution to bayesian 
tracking for non-linear, non-gaussian systems. 
%todo: might need to reference attempts to model a system of our type with
%unscented kalman filter
The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF can introduce bias; however, this effect can be
minimized by alterring the initial weights in the mixture pdf. Of course
every gradient descent must choose starting points and it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose a set or stream of measurements
are given, $\{y(t), t = 1, 2, 3, ... T\}$, where $T$ is permitted to go
 to infinity. Then the goal is to find the 
parameters, $\hat{\theta}$, and underlying state time series, $\hat{x}[0:T]$
that minimize the difference between $\hat{y}[0:T]$
and $y[0:T]$. In our case, we will assume that we know the form
of the model, which is based on first principals, and that
there is some true $\theta$ and a true time-series of underlying
state variable, $x[0:T]$ that drives $y[0:T]$. Assuming a model form 
such as we do here reduces model variance, potentially at the cost of increased
bias (or systematic) error. We will assume a basic state space model:

@@

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, the systems are assumed to be time invariant. This 
assumption is based on the idea that if you froze the system for $\Delta t$
seconds, when unfrozen the system would continue as if nothing happend. 
Few biological systems are predictible enough for them to be summarized
by a time varying function. Although the heart may seem like an obvious
exception, period between heartbeats vary often enough that prediction
would necessate another state-space model. In short, we
assume no parameters are time varying, because not enough information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, so unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Third, we will assume noise is additive.
Finally, $x(t)$ will encapsulate $\theta$, the unknown model constants, which
means that the vector $\dot{x}$ will always have members
that are 0. The results of these assumptions are a simplified version of the
state space equations:

@@

$\nu_x$ will result in something akin to adding a Weiner process to $y[0:T]$, and thus
will include low frequency noise. $\nu_y$ on the other hand will cause i.i.d. noise
in $y[0:T]$. For some of the tests, I will use de-trending methods to reduce the effects of 
$\nu_x$, the remainder of which should be re-factored into $\nu_y$. Both $\nu_x$ and $\nu_y$
have biological and non-biological sources. MR can lead to both types of noise, 
as demonstrated in \cite{drift}. Meanwhile changes in metabolism, heart rate, or
other biochemical intervention could all lead to either $\nu_x$ or $\nu_y$.

The goal of the particle filter is to evolve a probably distribution 
$Pr(\hat{x}(T) | u[0:T], y[0:T])$,
that asymptotically approaches the probability distribution $Pr(x(T) | u[0:T])$.
Considering that $y$ contains measurement noise and noise in $x$ can drive
changes in $y$, it is clear that $Pr(x(t) | u[0:T]$ is not a single true value
but a true posterior. 
To begin with, the particle filter starts with a proposal distribution, and $N_p$
particles need to be drawn from that distribution, $q(x)$:
\begin{equation}
\{[\hat{x}_i(0),w_i] : \hat{x}_i(0) \sim q(x), w_i = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}
\end{equation}

Where $N_p$ is the number of particles or points used to describe the prior 
using a Mixture PDF.

\begin{equation}
Pr(\hat{x}(0) = x(0)) = \sum_{i=1}^{N_p} w_i\delta(x(0) - \hat{x}_i(0) ) dx
\end{equation}

where

\begin{equation}
w_i(0) \propto \frac{p(x_i(0))}{q(x_i(0))}
\end{equation}

is the \emph{normalized} initial weights. Note that initially $p(x_i(0)$ is
an approximation of the proposal distribution, $q(x)$, thus by dividing
by $q(x_i(0))$, the bias effect of the proposal distribution is negated.
For the BOLD signal all the parameters have been studied and have relatively well
known mean and variance, so a prior bias is actually preferred, so in our
case we did not initially normalize the particle weights. Note however, that
when updating weights it is necessary to use this equation, but in that case
$q$ will not be the prior distribution but $p(\hat{x}(T) | \hat{x}(T-1)$

this the definition of the weight, so when weights are updated it is necessary
to divide by the proposal distributionto converge
to the true distribution it is necessary to divide by the proposal distribution.

Where $\delta(x-x_0)$ is 1 when $x = x_0$ (the Kronecker delta function).
For strictly positive parameters we used a gamma distribution,
whereas for parameters that could be negative, we used a Gaussian distribution. Both
distributions utilized standard deviations twice that found in previous studies.

Once the probability, $Pr(\hat{x}(T) | \hat{x}[0:T-1], y[0:T-1]$ has been found
(initially this is just Mixture approximating the prior since no measurements are 
available and no previous probabilities are available), its possible to approximate
the probability for short times between before a measurement is available, by shifting
the probability according the progression of the state equations. This is only 
an approximate, since integrating $\nu_x$ should increase uncertainty as more 
time without a measurement passes. 
\begin{equation}
Pr(\hat{x}(T+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i\delta\left(x - (\hat{x}_i(T) + \int_T^{T+\Delta} \dot{\hat{x}}_i(t) dt) \right) dx
\end{equation}

When a measurement becomes available it is incorporated into the probability,
which should eventually cause the probability to converge. 
Since the weights are the probability of of $x(t)$ being exactly $x_i(t)$, 
the weights are defined as
\begin{equation}
  d
\end{equation}
